---
title: "Introduction to ggadjustedforest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggadjustedforest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Motivation

When a study asks about the causal effect of a single **exposure** on an
outcome, the coefficients for adjustment covariates (confounders) are almost
never the quantities of interest.  Reporting them can even mislead readers,
because confounder coefficients are not identified under the causal model — they
absorb collider bias, mediation pathways, and other artefacts depending on the
causal structure (Hernán & Robins, *Causal Inference: What If*, 2020; Westreich
& Greenland, *Am J Epidemiol*, 2013).

STROBE guidelines (Vandenbroucke et al., 2007) and recent work on the
**estimand framework** (ICH E9(R1), 2019) both emphasise that reporting should
clearly distinguish the target quantity from nuisance parameters.

`ggadjustedforest` operationalises this principle: it fits the models you
specify but **only exposes the exposure coefficient** in both the plot and the
table, hiding confounder estimates by design.

---

## Basic usage — logistic regression

```{r logistic}
library(ggadjustedforest)

data(mtcars)
mtcars$am <- as.integer(mtcars$am)   # binary outcome: automatic (0) vs manual (1)

result <- gg_adjusted_forest(
  data       = mtcars,
  outcome    = "am",
  exposure   = "hp",
  covariates = c("wt", "cyl"),
  model_type = "logistic",
  title      = "Effect of Horsepower on Transmission Type"
)
result$table
```

The returned object has three components:

| Component | Contents |
|---|---|
| `$plot` | Combined forest plot + table (ggplot2 / patchwork) |
| `$table` | Numeric data frame |
| `$formatted_table` | Character table with formatted CI strings |

To render the plot:

```{r plot-logistic}
result$plot
```

---

## Cumulative adjustment

When you want to visualise how the effect estimate changes as confounders are
added sequentially — a common presentation in epidemiological reporting — use
`cumulative = TRUE`:

```{r cumulative}
result_cum <- gg_adjusted_forest(
  data       = mtcars,
  outcome    = "am",
  exposure   = "hp",
  covariates = c("wt", "cyl", "disp"),
  cumulative = TRUE,
  title      = "Cumulative adjustment: hp on transmission"
)
result_cum$formatted_table[, c("model", "formatted", "p.value")]
```

You can rename the rows with `cumulative_labels`:

```{r cumulative-labels}
labels <- c(
  "Unadjusted"         = "Crude",
  "+ wt"               = "Adjusted for weight",
  "+ wt + cyl"         = "Adjusted for weight + cylinders",
  "+ wt + cyl + disp"  = "Fully adjusted"
)

result_cum2 <- gg_adjusted_forest(
  data              = mtcars,
  outcome           = "am",
  exposure          = "hp",
  covariates        = c("wt", "cyl", "disp"),
  cumulative        = TRUE,
  cumulative_labels = labels,
  title             = "Cumulative adjustment with custom labels"
)
result_cum2$plot
```

---

## Cox proportional hazards regression

For time-to-event outcomes supply `model_type = "coxph"` along with `time_var`
and `event_var`:

```{r cox}
lung <- survival::lung
lung$status01 <- as.integer(lung$status == 2)
lung <- stats::na.omit(lung[, c("time", "status01", "age", "sex", "ph.ecog")])

result_cox <- gg_adjusted_forest(
  data       = lung,
  outcome    = "status01",
  exposure   = "age",
  covariates = c("sex", "ph.ecog"),
  model_type = "coxph",
  time_var   = "time",
  event_var  = "status01",
  title      = "Effect of Age on Survival (lung cancer)"
)
result_cox$plot
```

---

## Customising appearance

All the major aesthetic parameters are exposed:

```{r custom}
gg_adjusted_forest(
  data           = mtcars,
  outcome        = "am",
  exposure       = "hp",
  covariates     = "wt",
  model_type     = "logistic",
  color          = "#2166ac",
  point_size     = 5,
  point_shape    = 18,          # diamond
  vline_color    = "firebrick",
  vline_linetype = "dotted",
  x_breaks       = c(0.9, 1.0, 1.1),
  title          = "Custom aesthetics"
)$plot
```

For linear models the x-axis is on the original scale (not log):

```{r linear}
gg_adjusted_forest(
  data       = mtcars,
  outcome    = "mpg",
  exposure   = "hp",
  covariates = c("wt", "cyl"),
  model_type = "linear",
  title      = "Effect of Horsepower on Fuel Efficiency"
)$plot
```

---

## Extracting the table only

Use `forest_table()` when you only need the numbers:

```{r table-only}
forest_table(
  data       = mtcars,
  outcome    = "am",
  exposure   = "hp",
  covariates = c("wt", "cyl"),
  model_type = "logistic"
)
```

---

## Comparing multiple outcomes side-by-side

`ggadjustedforest` intentionally does not provide a built-in multi-outcome
wrapper. Each outcome deserves its own carefully specified model, and bundling
them into a single function call obscures that. Instead, fit each outcome
separately and panel the plots using
[`cowplot::plot_grid()`](https://wilkelab.org/cowplot/):

```{r multi-cowplot, warning=FALSE, fig.height=5, fig.width=10}
# install.packages("cowplot")  # if needed
library(cowplot)

data(mtcars)
mtcars$am <- as.integer(mtcars$am)
mtcars$vs <- as.integer(mtcars$vs)

# Extract $plot — cowplot takes ggplot2 objects, not the ggadjustedforest wrapper
p_am <- gg_adjusted_forest(
  data = mtcars, outcome = "am", exposure = "hp",
  covariates = c("wt", "cyl"), model_type = "logistic",
  title = "Transmission", show_table = FALSE
)$plot

p_vs <- gg_adjusted_forest(
  data = mtcars, outcome = "vs", exposure = "hp",
  covariates = c("wt", "cyl"), model_type = "logistic",
  title = "Engine Shape", show_table = FALSE
)$plot

cowplot::plot_grid(p_am, p_vs, ncol = 1, align = "v", axis = "lr")
```

This approach gives full control over each panel — different covariate sets,
model types, or axis scales per outcome — and composes cleanly with any other
ggplot2 object via `cowplot`. Use `align = "v"` with `axis = "lr"` to keep
the y-axis labels aligned across panels.

---

## References

- Hernán MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC.
- Vandenbroucke JP et al. (2007). Strengthening the Reporting of Observational
  Studies in Epidemiology (STROBE). *PLoS Med* 4(10): e297.
- Westreich D, Greenland S (2013). The table 2 fallacy: presenting and
  interpreting confounder and modifier coefficients. *Am J Epidemiol*
  177(4): 292–298.
- ICH E9(R1) (2019). Statistical Principles for Clinical Trials: Addendum on
  Estimands and Sensitivity Analysis in Clinical Trials. ICH Harmonised
  Guideline.
