---
title: "Introduction to ggadjustedforest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggadjustedforest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning    = FALSE
)
```

## Installation

```r
# Install from GitHub (CRAN submission pending)
# install.packages("remotes")
remotes::install_github("kriz98/gg_adjusted_forest")
```

## Motivation

When building multivariable models for causal inference, there is an **exposure**
of interest for which a causal estimand applies to. The coefficients for adjusted
covariates (confounders) can however be misinterpreted when presented together
with the estimand of interest. Reporting them can even mislead readers,
because confounder coefficients are not identified under the causal model, and are
susceptible to absorbing collider bias, mediation pathways, and other artefacts
depending on causal structures (Hernán & Robins, *Causal Inference: What If*, 2020; Westreich
& Greenland, *Am J Epidemiol*, 2013).

STROBE guidelines (Vandenbroucke et al., 2007) and recent work on the
**estimand framework** (ICH E9(R1), 2019) both emphasise that reporting should
clearly distinguish the target quantity from nuisance parameters.

`ggadjustedforest` operationalises this principle and fits the models you
specify but **only exposes the coefficient of interest** in both the plot and the
table, hiding confounder estimates by design. We also provide a cumulative
adjustment option, to show the effect of adjustment and model building during
exploratory phases of studies.

---

## Data

All examples use `colon_s` from the `finalfit` package — 929 patients with
colon cancer from the North Central Cancer Treatment Group (NCCTG) trial
(Moertel et al., *NEJM* 1990). The research question throughout is whether
having >4 positive lymph nodes (`node4`) increases the risk of 5-year mortality,
before and after adjusting for patient and tumour characteristics.

```{r setup}
library(ggadjustedforest)
library(finalfit)

data(colon_s)
colon_s$died_5yr <- as.integer(colon_s$mort_5yr == "Died")

confounders <- c("age", "sex.factor", "extent.factor", "differ.factor", "surg.factor")
```

---

## Basic usage — logistic regression

The pipe-friendly API puts `data` first so the function composes naturally with
`|>`:

```{r logistic}
result <- colon_s |>
  gg_adjusted_forest(
    outcome    = "died_5yr",
    exposure   = "node4",
    covariates = confounders,
    model_type = "logistic",
    title      = "Effect of Lymph Node Involvement (>4 nodes) on 5-Year Mortality"
  )
result$table
```

The returned object has three components:

| Component | Contents |
|---|---|
| `$plot` | Combined forest plot + table (ggplot2 / patchwork) |
| `$table` | Tibble of numeric estimates |
| `$formatted_table` | Tibble with formatted "OR (lower–upper)" strings |

To render the plot:

```{r plot-logistic}
result$plot
```

---

## Cumulative adjustment

When you want to visualise how the effect estimate changes as confounders are
added sequentially — a common presentation in epidemiological reporting — use
`cumulative = TRUE`. Use `cumulative_labels` to provide human-readable row names:

```{r cumulative, fig.height=5}
result_cum <- gg_adjusted_forest(
  data       = colon_s,
  outcome    = "died_5yr",
  exposure   = "node4",
  covariates = confounders,
  model_type = "logistic",
  cumulative = TRUE,
  cumulative_labels = c(
    "Unadjusted"                                                         = "Unadjusted",
    "+ age"                                                              = "+ Age",
    "+ age + sex.factor"                                                 = "+ Sex",
    "+ age + sex.factor + extent.factor"                                 = "+ Extent of spread",
    "+ age + sex.factor + extent.factor + differ.factor"                 = "+ Tumour differentiation",
    "+ age + sex.factor + extent.factor + differ.factor + surg.factor"   = "+ Time from surgery"
  ),
  title = "Cumulative Adjustment: Lymph Node Involvement on 5-Year Mortality"
)
result_cum$plot
```

The numeric table is available for downstream use:

```{r cumulative-table}
result_cum$formatted_table[, c("model", "formatted", "p.value")]
```

---

## Cox proportional hazards regression

For time-to-event outcomes supply `model_type = "coxph"` along with `time_var`
and `event_var`:

```{r cox}
result_cox <- gg_adjusted_forest(
  data       = colon_s,
  outcome    = "status",
  exposure   = "node4",
  covariates = confounders,
  model_type = "coxph",
  time_var   = "time.years",
  event_var  = "status",
  title      = "Hazard of Death by Lymph Node Involvement (Cox model)"
)
result_cox$plot
```

---

## Comparing multiple outcomes side-by-side

`ggadjustedforest` intentionally does not provide a built-in multi-outcome
wrapper. Each outcome deserves its own carefully specified model, and bundling
them into a single function call obscures that. Instead, fit each outcome
separately and stack the plots with `patchwork`, which is already a dependency
of `ggadjustedforest`:

```{r multi-patchwork, fig.height=6}
library(patchwork)

colon_s$died_all <- colon_s$status

p_5yr <- gg_adjusted_forest(
  data = colon_s, outcome = "died_5yr", exposure = "node4",
  covariates = confounders, model_type = "logistic",
  title = "5-Year Mortality", show_table = FALSE
)$plot

p_all <- gg_adjusted_forest(
  data = colon_s, outcome = "died_all", exposure = "node4",
  covariates = confounders, model_type = "logistic",
  title = "Death (all follow-up)", show_table = FALSE
)$plot

p_5yr / p_all
```

This approach gives full control over each panel — different covariate sets,
model types, or axis scales per outcome. The `/` operator stacks plots
vertically; use `|` for side-by-side.

---

## Customising appearance

All the major aesthetic parameters are exposed:

```{r custom}
gg_adjusted_forest(
  data           = colon_s,
  outcome        = "died_5yr",
  exposure       = "node4",
  covariates     = confounders,
  model_type     = "logistic",
  color          = "#2166ac",
  point_size     = 5,
  point_shape    = 18,
  vline_color    = "firebrick",
  vline_linetype = "dotted",
  title          = "Custom aesthetics"
)$plot
```

---

## Extracting the table only

Use `forest_table()` when you only need the numbers:

```{r table-only}
forest_table(
  data       = colon_s,
  outcome    = "died_5yr",
  exposure   = "node4",
  covariates = confounders,
  model_type = "logistic"
)
```

---

## References

- Hernán MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC.
- Moertel CG et al. (1990). Levamisole and fluorouracil for adjuvant therapy of
  resected colon carcinoma. *N Engl J Med* 322(6): 352–358.
- Vandenbroucke JP et al. (2007). Strengthening the Reporting of Observational
  Studies in Epidemiology (STROBE). *PLoS Med* 4(10): e297.
- Westreich D, Greenland S (2013). The table 2 fallacy: presenting and
  interpreting confounder and modifier coefficients. *Am J Epidemiol*
  177(4): 292–298.
- ICH E9(R1) (2019). Statistical Principles for Clinical Trials: Addendum on
  Estimands and Sensitivity Analysis in Clinical Trials. ICH Harmonised
  Guideline.
