---
title: "Introduction to ggadjustedforest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggadjustedforest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Installation

```r
# Install from GitHub (CRAN submission pending)
# install.packages("remotes")
remotes::install_github("kriz98/gg_adjusted_forest")
```

## Motivation

When building multivariable models for causal inference, there is an **exposure** 
of interest for which a causal estimand applies to. The coefficients for adjusted
covariates (confounders) can however be misinterpreted when presented together 
with the estimand of interest. Reporting them can even mislead readers,
because confounder coefficients are not identified under the causal model, and are
susceptible to absorbing collider bias, mediation pathways, and other artefacts 
depending on causal structures (Hernán & Robins, *Causal Inference: What If*, 2020; Westreich
& Greenland, *Am J Epidemiol*, 2013).

STROBE guidelines (Vandenbroucke et al., 2007) and recent work on the
**estimand framework** (ICH E9(R1), 2019) both emphasise that reporting should
clearly distinguish the target quantity from nuisance parameters.

`ggadjustedforest` operationalises this principle and fits the models you
specify but **only exposes the coefficient of interest** in both the plot and the
table, hiding confounder estimates by design. We also provide a cumulative 
adjustment option, to show the effect of adjustment and model building during
exploratory phases of studies. 

---

## Basic usage — logistic regression

```{r logistic}
library(ggadjustedforest)

data(mtcars)
mtcars$am <- as.integer(mtcars$am)   # binary outcome: automatic (0) vs manual (1)

result <- gg_adjusted_forest(
  data       = mtcars,
  outcome    = "am",
  exposure   = "hp",
  covariates = c("wt", "cyl"),
  model_type = "logistic",
  title      = "Effect of Horsepower on Transmission Type"
)
result$table
```

The returned object has three components:

| Component | Contents |
|---|---|
| `$plot` | Combined forest plot + table (ggplot2 / patchwork) |
| `$table` | Numeric data frame |
| `$formatted_table` | Character table with formatted CI strings |

To render the plot:

```{r plot-logistic}
result$plot
```

---

## Cumulative adjustment

When you want to visualise how the effect estimate changes as confounders are
added sequentially — a common presentation in epidemiological reporting — use
`cumulative = TRUE`:

```{r cumulative}
result_cum <- gg_adjusted_forest(
  data       = mtcars,
  outcome    = "am",
  exposure   = "hp",
  covariates = c("wt", "cyl", "disp"),
  cumulative = TRUE,
  title      = "Cumulative adjustment: hp on transmission"
)
result_cum$formatted_table[, c("model", "formatted", "p.value")]
```

You can rename the rows with `cumulative_labels`:

```{r cumulative-labels}
labels <- c(
  "Unadjusted"         = "Crude",
  "+ wt"               = "Adjusted for weight",
  "+ wt + cyl"         = "Adjusted for weight + cylinders",
  "+ wt + cyl + disp"  = "Fully adjusted"
)

result_cum2 <- gg_adjusted_forest(
  data              = mtcars,
  outcome           = "am",
  exposure          = "hp",
  covariates        = c("wt", "cyl", "disp"),
  cumulative        = TRUE,
  cumulative_labels = labels,
  title             = "Cumulative adjustment with custom labels"
)
result_cum2$plot
```

---

## Cox proportional hazards regression

For time-to-event outcomes supply `model_type = "coxph"` along with `time_var`
and `event_var`:

```{r cox}
lung <- survival::lung
lung$status01 <- as.integer(lung$status == 2)
lung <- stats::na.omit(lung[, c("time", "status01", "age", "sex", "ph.ecog")])

result_cox <- gg_adjusted_forest(
  data       = lung,
  outcome    = "status01",
  exposure   = "age",
  covariates = c("sex", "ph.ecog"),
  model_type = "coxph",
  time_var   = "time",
  event_var  = "status01",
  title      = "Effect of Age on Survival (lung cancer)"
)
result_cox$plot
```

---

## Customising appearance

All the major aesthetic parameters are exposed:

```{r custom}
gg_adjusted_forest(
  data           = mtcars,
  outcome        = "am",
  exposure       = "hp",
  covariates     = "wt",
  model_type     = "logistic",
  color          = "#2166ac",
  point_size     = 5,
  point_shape    = 18,          # diamond
  vline_color    = "firebrick",
  vline_linetype = "dotted",
  x_breaks       = c(0.9, 1.0, 1.1),
  title          = "Custom aesthetics"
)$plot
```

For linear models the x-axis is on the original scale (not log):

```{r linear}
gg_adjusted_forest(
  data       = mtcars,
  outcome    = "mpg",
  exposure   = "hp",
  covariates = c("wt", "cyl"),
  model_type = "linear",
  title      = "Effect of Horsepower on Fuel Efficiency"
)$plot
```

---

## Extracting the table only

Use `forest_table()` when you only need the numbers:

```{r table-only}
forest_table(
  data       = mtcars,
  outcome    = "am",
  exposure   = "hp",
  covariates = c("wt", "cyl"),
  model_type = "logistic"
)
```

---

## Comparing multiple outcomes side-by-side

`ggadjustedforest` intentionally does not provide a built-in multi-outcome
wrapper. Each outcome deserves its own carefully specified model, and bundling
them into a single function call obscures that. Instead, fit each outcome
separately and stack the plots with `patchwork`, which is already a dependency
of `ggadjustedforest`:

```{r multi-patchwork, warning=FALSE, fig.height=5, fig.width=10}
library(patchwork)

data(mtcars)
mtcars$am <- as.integer(mtcars$am)
mtcars$vs <- as.integer(mtcars$vs)

# Extract $plot — patchwork composes ggplot2 objects directly
p_am <- gg_adjusted_forest(
  data = mtcars, outcome = "am", exposure = "hp",
  covariates = c("wt", "cyl"), model_type = "logistic",
  title = "Transmission", show_table = FALSE
)$plot

p_vs <- gg_adjusted_forest(
  data = mtcars, outcome = "vs", exposure = "hp",
  covariates = c("wt", "cyl"), model_type = "logistic",
  title = "Engine Shape", show_table = FALSE
)$plot

p_am / p_vs
```

This approach gives full control over each panel — different covariate sets,
model types, or axis scales per outcome. The `/` operator stacks plots
vertically; use `|` for side-by-side. Pass `plot_layout(guides = "collect")`
to share a legend if needed.

---

## References

- Hernán MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC.
- Vandenbroucke JP et al. (2007). Strengthening the Reporting of Observational
  Studies in Epidemiology (STROBE). *PLoS Med* 4(10): e297.
- Westreich D, Greenland S (2013). The table 2 fallacy: presenting and
  interpreting confounder and modifier coefficients. *Am J Epidemiol*
  177(4): 292–298.
- ICH E9(R1) (2019). Statistical Principles for Clinical Trials: Addendum on
  Estimands and Sensitivity Analysis in Clinical Trials. ICH Harmonised
  Guideline.
